Deep Recurrent Neural Networks for Supernovae Classification

### 2016-06-23 paper/sn.tex
arXiv ready
Paper

\begin{figure}
\centering
\includegraphics[width=88mm, angle=0]{network.pdf}
\caption{\label{fig:network} Bidirectional recurrent neural network for sequence classification. The input vectors at each sequential step are fed into a pair of bidirectional hidden layers, which can propagate information forwards and backwards. These are then merged to obtain a consensus view of the network, and finally a softmax layer computes classification probabilities. 
 }
\end{figure}

{\em Recurrent neural networks} (RNNs) are a class of artificial neural network  that can learn about sequential data. They are commonly used for tasks such speech recognition and language translation, but have several possible applications in astronomy  and cosmology for processing temporal or spatial sequential data. RNNs have several properties which makes them suitable for sequential information. The inputs to the network are flexible, and they are able to recognise patterns with noisy data (for example the context of a word in a sentence relative to others can vary).

The main problem with vanilla RNNs  is that they are unable to store long-term information, so inputs at the end of a sequence have no knowledge of inputs at the start. Several types of RNNs have been proposed to solve this problem, including {\em Long Short-Term Memory} (LSTM)~\cite{LSTM} and {\em Gated Recurrent Unit} (GRU)~\cite{2014arXiv1412.3555C} units. These are similar in concept, in that information is able to flow through the network via a gating mechanism. One further problem with RNNs is that information is only able to flow in one direction. In {\em bidirectional} RNNs information is able to pass both forwards and backwards. Bidirectional LSTM networks have been shown to be particularly powerful for supervised sequence labelling, where sequential data is accompanied by a set of discrete labels.

The architecture of a typical bidirectional RNN for sequence labelling is shown in Fig.~\ref{fig:network}, where the squares represent {\rm neurons}. In this case the inputs, which are vectors at each sequential step, are connected to two hidden RNN layers, which can be either vanilla RNN or memory units.  Each hidden layer contains a number of hidden units, and in each layer information flows either forwards or backwards,  but no information passes between the two directions. Several hidden layers can be stacked to form {\em deep} neural networks. Deep networks are capable of learning higher-level temporal or spatial representations, and complex relationships between the inputs and outputs.

@@

 In the network, each neuron is connected to another by a weight matrix, and the optimal weights are found by back propagating the errors from a {\em loss function} of the output layer. For classification problems, this is typically the categorical cross-entropy between predictions and targets, defined as 
\begin{equation}
L= -\sum_{i,j} t_{i,j} \log \left( p_{i,j} \right)
\end{equation}
where $i,j$ run over the class labels, $t_{i,j}$ are the targets for each class (either 0 or 1) and $p_{i,j}$ are the predicted probabilities.

@@

 Observations are then of the form in Tab.~\ref{tab:augment}, where any missing values are denoted by a dash. In order to impute the missing value of $i$, we use {\em data augmentation} and randomly select a value between $i_1$ or $i_3$. We make 5 random augmentations of all missing data, thereby increasing the size of the dataset by a factor of 5. We can test the importance of augmentation by training each augmentation separately and  comparing the change in accuracy. We find that augmentation gives a percent level improvement in accuracy.

@@

\begin{figure}
\centering
\includegraphics[width=88mm, angle=0]{DES_SN551675.pdf}
\caption{\label{fig:lightcurve} (Top) Example light curve in the 4 $g, r, i, z$ bands for SN ID 551675 (a type Ia) in the Supernovae Photometric Classification Challenge data~\cite{Kessler:2010wk}. The data has been processed using augmentation so there is a $g, r, i, z$ value at each sequential step. (Bottom) Type Ia probability as a function of time from a 2 layer LSTM model, trained with around $10^4$ supernovae and SN 551675 excluded. The final probability gives $99.5\%$ confidence that the supernovae is of type Ia. 
 }
\end{figure}

The data comes in two types, those with and those without the  host galaxy photometric redshift. Each dataset is split into a training and test set, with the training set containing a spectroscopically confirmed supernovae type and redshift. It is important that augmented data with the same supernovae ID go into either the training {\em or} test set, not a mix of both. The original SPCC data consisted of 1103 training samples. The answer keys were subsequently made available for the test set~\cite{Kessler:2010qj}.

@@

\begin{figure}
\centering
\includegraphics[width=88mm, angle=0]{lstm_dropout_05_loss.pdf}
\caption{\label{fig:loss} (Left) Training loss (green) vs test loss (blue) for a unidirectional 2 layer  LSTM network with 16 hidden units in each layer. (Right) Training accuracy (green) vs test accuracy (blue) for the same network.  }
\end{figure}

@@

We also test the same model using a training fraction of 0.25, around 5000 supernovae, close to the lower  end of the number likely to be followed up for the LSST. Again, after 5 randomised runs and training for 200 epochs the accuracy was \comment{$***\pm***$}, the AUC was \comment{$***\pm***$} and \comment{$F_1 =***\pm***$}. This suggests that the classifer metrics scale with the data \comment{$\propto***$}.

For $5.2\%$ of the representative SPCC data, the training dataset is so small that over-fitting is more severe. Using the same 2 layer LSTM network with 16 hidden units and a dropout of 0.5 we find a notable increase in the test loss after $\sim 20$ epochs, but the accuracy and other metrics remain relatively constant ($F_1$ values, for example, of 0.35 to 0.4 were obtained). The reason for this apparent discrepancy is that the accuracy, say, simply takes the maximum value of the softmax output layer. For example, a 2 class problem with output probabilities [0.6, 0.4] and target [1, 0] has the same accuracy as one with output probabilities [0.8, 0.2]. The loss in the latter case would be lower however, and represents increased confidence of the network in its predictions. We therefore reject models with over-fitting and an increasing cross-entropy loss, at the expense of metrics such as $F_1$, and decrease the model complexity.

@@

There are several possibilities for future work. One of the advantages of recurrent neural networks is that inputs are agnostic, so the impact of any additional inputs could be explored. It would be possible, for example, to even pass the raw images in each filter though a convolutional network and use those as inputs. We have considered a representative training sample, but spectroscopic follow up surveys may be biased. The performance of the network could be measured against selection bias, and the results used to inform the best follow up strategy.  Further work could also be performed to optimise the early detection probability of the network.  Finally, to improve performance in the small data regime one can use {\em transfer learning.} Here, a more complex network is pre-trained on simulations or existing data from other surveys then the weights of the network are fine-tuned on the new, smaller dataset. The simulated SPCC data used in this work are based on the DES instrument, so it would be interesting to apply transfer learning to real DES data.

>>>>

\begin{figure}
\centering
\includegraphics[width=88mm, angle=0]{1.pdf}
\caption{\label{fig:network} Bidirectional recurrent neural network for sequence classification. The input vectors at each sequential step are fed into a pair of bidirectional hidden layers, which can propagate information forwards and backwards. These are then merged to obtain a consensus view of the network, and finally a softmax layer computes classification probabilities. 
 }
\end{figure}

{\em Recurrent neural networks} (RNNs) are a class of artificial neural network  that can learn about sequential data. They are commonly used for tasks such speech recognition and language translation, but have several possible applications in astronomy  and cosmology for processing temporal or spatial sequential data. RNNs have several properties which makes them suitable for sequential information. The inputs to the network are flexible, and they are able to recognise patterns with noisy data (for example the context of a word in a sentence relative to others can vary, or a time stream can contain instrument noise).

The main problem with vanilla RNNs  is that they are unable to store long-term information, so inputs at the end of a sequence have no knowledge of inputs at the start. This is a problem if the data has long-term correlations. Several types of RNNs have been proposed to solve this problem, including {\em Long Short-Term Memory} (LSTM)~\cite{LSTM} and {\em Gated Recurrent Unit} (GRU)~\cite{2014arXiv1412.3555C} units. These are similar in concept, in that information is able to flow through the network via a gating mechanism. One further problem with RNNs is that information is only able to flow in one direction. In {\em bidirectional} RNNs information is able to pass both forwards and backwards. Bidirectional LSTM networks have been shown to be particularly powerful for supervised sequence labelling, where sequential data is accompanied by a set of discrete labels.

The architecture of a typical bidirectional RNN for sequence labelling is shown in Fig.~\ref{fig:network}, where the squares represent {\rm neurons}. In this case the inputs, which are vectors at each sequential step, are connected to two hidden RNN layers, which can be either vanilla RNN or memory units.  Each hidden layer contains a number of hidden units (capable of storing information), and in each layer information flows either forwards or backwards,  but no information passes between the two directions. Several hidden layers can be stacked to form {\em deep} neural networks. Deep networks are capable of learning higher-level temporal or spatial representations, and complex relationships between the inputs and outputs.

@@

 In the network, each neuron is connected to another by a weight matrix, and the optimal weights are found by back-propagating the errors from a {\em loss function} of the output layer. For classification problems, this is typically the categorical cross-entropy between predictions and targets, defined as 
\begin{equation}
L= -\sum_{i,j} t_{i,j} \log \left( p_{i,j} \right)
\end{equation}
where $i,j$ run over the class labels, $t_{i,j}$ are the targets for each class (either 0 or 1) and $p_{i,j}$ are the predicted probabilities. Back-propogation essentially takes the derivative of the loss with respect to the weights $W$ of the output layer, $\partial L/\partial W$, and uses the chain rule to update the weights in the network.

@@

 Observations are then of the form in Tab.~\ref{tab:augment}, where any missing values are denoted by a dash. In order to impute the missing value of $i$, we use {\em data augmentation} and randomly select a value between $i_1$ or $i_3$. We make 5 random augmentations of all missing data, thereby increasing the size of the dataset by a factor of 5. We can test the importance of this by training each augmentation separately and  comparing the change in accuracy, which we find is $\sim 1\%$. Training with multiple augmentations at once gives the best performance, since the network will learn  to ignore random filled values.

@@

\begin{figure}
\centering
\includegraphics[width=88mm, angle=0]{2.pdf}
\caption{\label{fig:lightcurve} (Top) Example light curve in the 4 $g, r, i, z$ bands for SN ID 551675 (a type Ia) in the Supernovae Photometric Classification Challenge data~\cite{Kessler:2010wk}. The data has been processed using augmentation so there is a $g, r, i, z$ value at each sequential step. (Bottom) Type Ia probability as a function of time from a 2 layer LSTM model, trained with around $10^4$ supernovae and SN 551675 excluded. The final probability gives $99.5\%$ confidence that the supernovae is of type Ia. 
 }
\end{figure}

The data comes in two types, those with and those without the  host galaxy photometric redshift. Each dataset is split into a training and test set, with the training set containing a spectroscopically confirmed supernovae type and redshift. It is important that augmented data with the same supernovae ID go into either the training {\em or} test set, not a mix of both, otherwise the training and test data will not be independent. The original SPCC data consisted of 1103 training samples. The answer keys were subsequently made available for the test set~\cite{Kessler:2010qj}.

@@

\begin{figure}
\centering
\includegraphics[width=88mm, angle=0]{3.pdf}
\caption{\label{fig:loss} (Left) Training loss (green) vs test loss (blue) for a unidirectional 2 layer  LSTM network with 16 hidden units in each layer. (Right) Training accuracy (green) vs test accuracy (blue) for the same network.  }
\end{figure}

@@

We also test the same model using a training fraction of 0.25 (around 5000 supernovae), closer to the lower  end of the number likely to be followed up for the LSST. Again, after 5 randomised runs and training for 200 epochs we obtain an accuracy of $93.3 \pm 0.4\%$, an AUC of $0.978 \pm 0.002$ and $F_1 = 0.57 \pm 0.02$. The corresponding type Ia purity and completeness are 85\% and 87\% respectively. The $F_1$ metric, for example, has degraded by $\sim 10\%$ for a reduction in data of $50\%$.

For $5.2\%$ of the representative SPCC data, the training dataset is so small that over-fitting is more severe. Using the same 2 layer LSTM network with 16 hidden units and a dropout of 0.5 we find a notable increase in the test loss after $\sim 20$ epochs, but the accuracy and other metrics remain relatively constant ($F_1$ values, for example, of 0.35 to 0.4 were obtained). The reason for this apparent discrepancy is that the accuracy, say, simply takes the maximum value of the softmax output layer. For example, a 2 class problem with output probabilities [0.6, 0.4] and target [1, 0] has the same accuracy as one with output probabilities [0.8, 0.2]. The loss in the latter case would be lower however, and represents increased confidence of the network in its predictions. We therefore reject models with severe over-fitting and an increasing cross-entropy loss, at the expense of metrics such as $F_1$, and decrease the model complexity.

@@

There are several possibilities for future work. One of the advantages of recurrent neural networks is that inputs are agnostic, so the impact of any additional inputs could be explored. It would be possible, for example, to even pass the raw images in each filter though a convolutional network and use those as inputs. We have considered a representative training sample, but spectroscopic follow up surveys may be biased. The performance of the network could be measured against selection bias, and the results used to inform the best follow up strategy.  Further work could also be performed to optimise the early detection probability of the network.  Finally, to improve performance in the small data regime one can use {\em transfer learning.} Here, a more complex network is pre-trained on simulations or existing data from other surveys, then the weights of the network are fine-tuned on the new, smaller dataset. The simulated SPCC data used in this work are based on the DES instrument, so it would be interesting to apply transfer learning to real DES data.
